--
title: "Case Study: Spam Email Prediction"
author: "Jaclyn Coate, Megan Riley, Reagan Meagher"
date: "2/13/2021"
output: pdf_document
---

# Introduction

The digital world is something that has increasingly changed the way organizations and businesses communicate with their consumers. As digital memory becomes more cost effective it takes no effort at all for a company to render and send large volumes of emails in an attempt to engage their end users. While is a positive shift for companies in their effort to grow their business, consumers are inundated with spam emails that are problematic for those who use their email for other purposes than just shopping. Individuals can easily spot their spam emails and skip or delete them. But this manual effort can be strenuous and take a long time based on sheer volume of spam emails alone. This may not be a big issue for the casual email users who only use it for online shopping and occasional correspondence. But the individuals who live in the digital world would much rather not spend their time sorting email, but rather leveraging their email for their end goals.

This leads us to the case study we present today; below are our efforts to make sure we have a strong spam email filter to take the burden off of email users for manually filtering their spam emails out. The first question is how to apply a filter that effectively filters spam emails. Our process for an automatic procedure closely resembles the process a human might go through in order to determine if their email is spam. We review the _subject line_ and if we can’t tell from that we examine the _body_ without ever opening an _attachment_ in case of a virus. With cleaning and examining these different elements of an email programmatically we can leverage extremely powerful machine learning algorithmic classifiers such as: _NaiveBayes_, _Decisions Tree Classifier_, and _XGBoost_/_Support Vector Machine_.

```{r environment set up, echo=FALSE, results='markup', warning=FALSE}
#install.packages("tm")
library(tm)
#install.packages("naivebayes")
library(naivebayes)
#install.packages("e1071")
library(e1071)
#install.packages("rpart")
library(rpart)
#install.packages("randomForest")
library(randomForest)
#install.packages("xgboost")
library(xgboost)
#install.packages("caret")
library(caret)
Sys.setlocale("LC_ALL", "pt_PT.ISO8859-1")
```

Below is a full breakdown of functions levereaged in order to grab, split, identify boundaries, and many more cleaning requiremetns in order to processing the lagnagues contained within email messages/

```{r }
splitMessage = function(msg) {
  splitPoint = match("", msg)
  header = msg[1:(splitPoint-1)]
  body = msg[ -(1:splitPoint) ]
  return(list(header = header, body = body))
}
getBoundary = function(header) {
  boundaryIdx = grep("boundary=", header)
  boundary = gsub('"', "", header[boundaryIdx])
  gsub(".*boundary= *([^;]*);?.*", "\\1", boundary)
}
dropAttach = function(body, boundary){
  
  bString = paste("--", boundary, sep = "")
  bStringLocs = which(bString == body)
  
  if (length(bStringLocs) <= 1) return(body)
  
  eString = paste("--", boundary, "--", sep = "")
  eStringLoc = which(eString == body)
  if (length(eStringLoc) == 0) 
    return(body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1)])
  
  n = length(body)
  if (eStringLoc < n) 
     return( body[ c( (bStringLocs[1] + 1) : (bStringLocs[2] - 1), 
                    ( (eStringLoc + 1) : n )) ] )
  
  return( body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1) ])
}
findMsgWords = 
function(msg, stopWords) {
 if(is.null(msg))
  return(character())
 words = unique(unlist(strsplit(cleanText(msg), "[[:blank:]\t]+")))
 
 # drop empty and 1 letter words
 words = words[ nchar(words) > 1]
 words = words[ !( words %in% stopWords) ]
 invisible(words)
}
cleanText =
function(msg)   {
  tolower(gsub("[[:punct:]0-9[:space:][:blank:]]+", " ", msg))
}
processAllWords = function(dirName, stopWords)
{
       # read all files in the directory
  fileNames = list.files(dirName, full.names = TRUE)
       # drop files that are not email, i.e., cmds
  notEmail = grep("cmds$", fileNames)
  if ( length(notEmail) > 0) fileNames = fileNames[ - notEmail ]
  messages = lapply(fileNames, readLines, encoding = "latin1")
  
       # split header and body
  emailSplit = lapply(messages, splitMessage)
       # put body and header in own lists
  bodyList = lapply(emailSplit, function(msg) msg$body)
  headerList = lapply(emailSplit, function(msg) msg$header)
  rm(emailSplit)
  
       # determine which messages have attachments
  hasAttach = sapply(headerList, function(header) {
    CTloc = grep("Content-Type", header)
    if (length(CTloc) == 0) return(0)
    multi = grep("multi", tolower(header[CTloc])) 
    if (length(multi) == 0) return(0)
    multi
  })
  
  hasAttach = which(hasAttach > 0)
  
       # find boundary strings for messages with attachments
  boundaries = sapply(headerList[hasAttach], getBoundary)
  
       # drop attachments from message body
  bodyList[hasAttach] = mapply(dropAttach, bodyList[hasAttach], 
                               boundaries, SIMPLIFY = FALSE)
  
       # extract words from body
  #ERROR
  msgWordsList = lapply(bodyList, findMsgWords, stopWords)
  
  invisible(msgWordsList)
}
```

```{r}
#Directory paths
spamPath = "/Users/Jaco/Desktop/SMU/DS_7333_Quantifying_the_World/CaseStudies/Unit6_CaseStudy3/SpamData"
dirNames = list.files(path = paste(spamPath, sep = .Platform$file.sep))
fullDirNames = paste(spamPath, dirNames, sep = .Platform$file.sep)
#Set up Stop Words
stopWords = stopwords()
cleanSW = tolower(gsub("[[:punct:]0-9[:blank:]]+", " ", stopWords))
SWords = unlist(strsplit(cleanSW, "[[:blank:]]+"))
SWords = SWords[ nchar(SWords) > 1 ]
stopWords = unique(SWords)
```

```{r, warning=FALSE, echo=F}
# Grabbing all valide words from all emails
msgWordsList = lapply(fullDirNames, processAllWords, stopWords = stopWords)
numMsgs = sapply(msgWordsList, length)
numMsgs
```

```{r}
#Below we are adding the true labels on if messagse are spam or not spam. This is our known determination because it was hand labeled.
isSpam = rep(c(FALSE, FALSE, FALSE, TRUE, TRUE), numMsgs)
msgWordsList = unlist(msgWordsList, recursive = FALSE)
```

Below we are building an index of messages and performing a simple Bayes classification of the messages.

```{r, echo=F, include=F}
numEmail = length(isSpam)
numSpam = sum(isSpam)
numHam = numEmail - numSpam

set.seed(418910)
#Test / Train Split
testSpamIdx = sample(numSpam, size = floor(numSpam/3))
testHamIdx = sample(numHam, size = floor(numHam/3))
#Split out into test and train data
testMsgWords = c((msgWordsList[isSpam])[testSpamIdx],
                 (msgWordsList[!isSpam])[testHamIdx] )
trainMsgWords = c((msgWordsList[isSpam])[ - testSpamIdx], 
                  (msgWordsList[!isSpam])[ - testHamIdx])
testIsSpam = rep(c(TRUE, FALSE), 
                 c(length(testSpamIdx), length(testHamIdx)))
trainIsSpam = rep(c(TRUE, FALSE), 
                 c(numSpam - length(testSpamIdx), 
                   numHam - length(testHamIdx)))
```

Probability esitmation on the Training Data. HOW CAN WE DISPLAY THIS?

```{r, echo=F, include=F}
# the number of unique words in the training data
bow = unique(unlist(trainMsgWords))
length(bow)
# the number of occurances of the unique words in a spam message
spamWordCounts = rep(0, length(bow))
names(spamWordCounts) = bow
tmp = lapply(trainMsgWords[trainIsSpam], unique)
tt = table( unlist(tmp) )
spamWordCounts[ names(tt) ] = tt

# number of occurrences of spam and non-spam for each word on the list
computeFreqs =
function(wordsList, spam, bow = unique(unlist(wordsList)))
{
   # create a matrix for spam, ham, and log odds
  wordTable = matrix(0.5, nrow = 4, ncol = length(bow), 
                     dimnames = list(c("spam", "ham", 
                                        "presentLogOdds", 
                                        "absentLogOdds"),  bow))
   # For each spam message, add 1 to counts for words in message
  counts.spam = table(unlist(lapply(wordsList[spam], unique)))
  wordTable["spam", names(counts.spam)] = counts.spam + .5
   # Similarly for ham messages
  counts.ham = table(unlist(lapply(wordsList[!spam], unique)))  
  wordTable["ham", names(counts.ham)] = counts.ham + .5  
   # Find the total number of spam and ham
  numSpam = sum(spam)
  numHam = length(spam) - numSpam
   # Prob(word|spam) and Prob(word | ham)
  wordTable["spam", ] = wordTable["spam", ]/(numSpam + .5)
  wordTable["ham", ] = wordTable["ham", ]/(numHam + .5)
  
   # log odds
  wordTable["presentLogOdds", ] = 
     log(wordTable["spam",]) - log(wordTable["ham", ])
  wordTable["absentLogOdds", ] = 
     log((1 - wordTable["spam", ])) - log((1 -wordTable["ham", ]))
  invisible(wordTable)
}

# applying spam versus non-spam function to training data
trainTable = computeFreqs(trainMsgWords, trainIsSpam)
```

Below we are creating a classification test dataset. There is potential for no words in the training data. Below we are claculating the log odds without new words.

```{r, echo=F}
# Note: the first message is spac, and the log likelihood is positive
newMsg = testMsgWords[[1]]
newMsg = newMsg[!is.na(match(newMsg, colnames(trainTable)))]
present = colnames(trainTable) %in% newMsg
sum(trainTable["presentLogOdds", present]) + 
  sum(trainTable["absentLogOdds", !present])
# Note: A negative value with a large log likelihood ratio indiciates a non-spam message
newMsg = testMsgWords[[ which(!testIsSpam)[1] ]]
newMsg = newMsg[!is.na(match(newMsg, colnames(trainTable)))]
present = (colnames(trainTable) %in% newMsg)
sum(trainTable["presentLogOdds", present]) + 
     sum(trainTable["absentLogOdds", !present])

# function to get log likelihood
computeMsgLLR = function(words, freqTable) 
{
       # Discards words not in training data.
  words = words[!is.na(match(words, colnames(freqTable)))]
       # Find which words are present
  present = colnames(freqTable) %in% words
  sum(freqTable["presentLogOdds", present]) +
    sum(freqTable["absentLogOdds", !present])
}
```

Below we have displayed the distributions of the log likelihoods for each of the different classes: _spam_ versus _non-spam_. Due to the distribution spread of the Spam email messages being nearest zero (including the mean) we can state that the data shows our probability is more accurate at identifying the Spam versus the Non-spam messages. We also noted that below the quartile ranges of the _non-spam_ emails is much smaller than the spam emails. This tells us that the repeatability of results for identifying _non-spam_ messages is higher than the Spam messages. Simply stated this indicates, the words unique words identified in _non-spam_ messages are not as unique to _non-spam_ messages as those words used in Spam messages. Additionally, the words most used in _non-spam_ messages can also appear in _spam messages.

Analyzing from a logical perspective we can note that _spam_ emails often contain words that might be difficult for detectors to pick up, such as repetitive letters to change the n_grams of the words and _leet speak_. _Leet speak_ is the practice of replacing words with letters and such as an _E_ becomes a _3_ or an _0_ becomse a _0_.

Since the Naive Bayes approach cannot easily classify these nuanced characteristic we perform a _decision tree classifier_ and an _XGBoost_ analysis to compare results.


```{r}
# do not run twice
testLLR = sapply(testMsgWords, computeMsgLLR, trainTable)
tapply(testLLR, testIsSpam, summary)
# boxplots of log likelihoods
#pdf("SP_Boxplot.pdf", width = 6, height = 6)
spamLab = c("ham", "spam")[1 + testIsSpam]
boxplot(testLLR ~ spamLab, 
        ylab = "Log Likelihood Ratio",
       main = "Log Likelihood Ratio for Randomly Chosen Test Messages",
        ylim=c(-500, 500))
#dev.off()
```

Below we have Naive Bayes misclassifications and are calculating our Type I error rate with a function.

```{r}
typeIErrorRate = 
function(tau, llrVals, spam)
{
  classify = llrVals > tau
  sum(classify & !spam)/sum(!spam)
}
#tau 0
typeIErrorRate(0, testLLR,testIsSpam)
#tau -20
typeIErrorRate(-20, testLLR,testIsSpam)
typeIErrorRates = 
function(llrVals, isSpam) 
{
  o = order(llrVals)
  llrVals =  llrVals[o]
  isSpam = isSpam[o]
  idx = which(!isSpam)
  N = length(idx)
  list(error = (N:1)/N, values = llrVals[idx])
}
```

Below we have Naive Bayes misclassifications and are calculating our Type II error rate with a function.

```{r}
typeIIErrorRates = function(llrVals, isSpam) {
    
  o = order(llrVals)
  llrVals =  llrVals[o]
  isSpam = isSpam[o]
    
    
  idx = which(isSpam)
  N = length(idx)
  list(error = (1:(N))/N, values = llrVals[idx])
  }  
xI = typeIErrorRates(testLLR, testIsSpam)
xII = typeIIErrorRates(testLLR, testIsSpam)
tau01 = round(min(xI$values[xI$error <= 0.01]))
t2 = max(xII$error[ xII$values < tau01 ])
```

### Type I versus Type II Error Rates and Log-Likelihood Thresholds

Type I error rate is also known as a false positive error rate and occurs when the predictor incorrectly rejects a true null hypothesis. In our research the null hypothesis states that a a message is not spam. Below, in Figure XX, we have displayed the distribution of the Type I and Type II errors of our log likelihood values. Type I error rate is also known as a false positive error rate and occurs when the predictor incorrectly rejects a true null hypothesis. Type II error rate is also known as a false negative error rate and occurs when the predictor incorrectly accepts a false null hypothesis. As indicated by the orange marker, the threshold for determining spam is a log likelihood value of -41. Otherwise stated, any message that has a log likelihood value that is greater than -41 would be classified as a spam email and any message that has a log-likelihood of less than -41 is not a spam email.

```{r}
#pdf("LinePlotTypeI+IIErrors.pdf", width = 8, height = 6)
library(RColorBrewer)
cols = brewer.pal(9, "Set1")[c(3, 4, 5)]
plot(xII$error ~ xII$values,  type = "l", col = cols[1], lwd = 3,
     xlim = c(-300, 250), ylim = c(0, 1),
     xlab = "Log Likelihood Ratio Values", ylab="Error Rate", main = "Type I & Type II Error Rates: Spam vs Non-spam Log Likelihood Ratios")
points(xI$error ~ xI$values, type = "l", col = cols[2], lwd = 3)
legend(x = 50, y = 0.4, fill = c(cols[2], cols[1]),
       legend = c("Classify Ham as Spam", 
                  "Classify Spam as Ham"), cex = 0.8,
       bty = "n")
abline(h=0.01, col ="grey", lwd = 3, lty = 2)
text(-250, 0.05, pos = 4, "Type I Error = 0.01", col = cols[2])
mtext(tau01, side = 1, line = 0.5, at = tau01, col = cols[3])
segments(x0 = tau01, y0 = -.50, x1 = tau01, y1 = t2, 
         lwd = 2, col = "grey")
text(tau01 + 20, 0.05, pos = 4,
     paste("Type II Error = ", round(t2, digits = 2)), 
     col = cols[1])
```

NOTE: In the above figure the Type I error threshold is 0.01 and the threshold for Type II error rate is 0.04.

```{r}
#Below we are obtaining a tau so that one type of error is withint 1% using our 5-fold cross-validation.
#dev.off()
k = 5
numTrain = length(trainMsgWords)
partK = sample(numTrain)
tot = k * floor(numTrain/k)
partK = matrix(partK[1:tot], ncol = k)
testFoldOdds = NULL
for (i in 1:k) {
  foldIdx = partK[ , i]
  trainTabFold = computeFreqs(trainMsgWords[-foldIdx], trainIsSpam[-foldIdx])
  testFoldOdds = c(testFoldOdds, 
               sapply(trainMsgWords[ foldIdx ], computeMsgLLR, trainTabFold))
}
testFoldSpam = NULL
for (i in 1:k) {
  foldIdx = partK[ , i]
  testFoldSpam = c(testFoldSpam, trainIsSpam[foldIdx])
}
xFoldI = typeIErrorRates(testFoldOdds, testFoldSpam)
xFoldII = typeIIErrorRates(testFoldOdds, testFoldSpam)
tauFoldI = round(min(xFoldI$values[xFoldI$error <= 0.01]))
tFold2 = xFoldII$error[ xFoldII$values < tauFoldI ]
```






I CANNOT GET THIS CODE TO RUN AM RECEIVING THE ERRORS WE INIATIALLY SKIPPED BECAUSE OF THE ERRORS


```{r}
fullDirNames = paste(spamPath, "data", dirNames, 
                     sep = .Platform$file.sep)

indx = c(1:5, 15, 27, 68, 69, 329, 404, 427, 516, 852, 971)
fn = list.files(fullDirNames[1], full.names = TRUE)[indx]
sampleEmail = sapply(fn, readLines)        

msg = sampleEmail[[1]]

smallNums = rep((1/2)^40, 2000000)
largeNum = 10000
print(sum(smallNums), digits = 20)
print(largeNum + sum(smallNums), digits = 20)
for (i in 1:length(smallNums)) {
  largeNum = largeNum + smallNums[i]
}
print(largeNum, digits = 20)
sampleSplit = lapply(sampleEmail, splitMessage)
header = sampleSplit[[1]]$header
header[1:12]
header[1] = sub("^From", "Top-From:", header[1])
header[1]
headerPieces = read.dcf(textConnection(header), all = TRUE)
headerPieces[, "Delivered-To"]
headerVec = unlist(headerPieces)
dupKeys = sapply(headerPieces, function(x) length(unlist(x)))
names(headerVec) = rep(colnames(headerPieces), dupKeys)
headerVec[ which(names(headerVec) == "Delivered-To") ]
length(headerVec)
length(unique(names(headerVec)))
```

```{r}
# Additional functions for???
processHeader = function(header)
{
       # modify the first line to create a key:value pair
  header[1] = sub("^From", "Top-From:", header[1])
  
  headerMat = read.dcf(textConnection(header), all = TRUE)
  headerVec = unlist(headerMat)
  
  dupKeys = sapply(headerMat, function(x) length(unlist(x)))
  names(headerVec) = rep(colnames(headerMat), dupKeys)
  
  return(headerVec)
}
headerList = lapply(sampleSplit, 
                    function(msg) {
                      processHeader(msg$header)} )
contentTypes = sapply(headerList, function(header) 
                                        header["Content-Type"])
names(contentTypes) = NULL
contentTypes
hasAttach = grep("^ *multi", tolower(contentTypes))
hasAttach
boundaries = getBoundary(contentTypes[ hasAttach ])
boundaries
boundary = boundaries[9]
body = sampleSplit[[15]]$body
bString = paste("--", boundary, sep = "")
bStringLocs = which(bString == body)
bStringLocs
eString = paste("--", boundary, "--", sep = "")
eStringLoc = which(eString == body)
eStringLoc
diff(c(bStringLocs[-1], eStringLoc))
```

```{r}
### This code has mistakes in it - and we fix them later!
processAttach = function(body, contentType){
  
  boundary = getBoundary(contentType)
  
  bString = paste("--", boundary, "$", sep = "")
  bStringLocs = grep(bString, body)
  
  eString = paste("--", boundary, "--$", sep = "")
  eStringLoc = grep(eString, body)
  
  n = length(body)
  
  if (length(eStringLoc) == 0) eStringLoc = n + 1
  if (length(bStringLocs) == 1) attachLocs = NULL
  else attachLocs = c(bStringLocs[-1],  eStringLoc)
  
  msg = body[ (bStringLocs[1] + 1) : min(n, (bStringLocs[2] - 1), 
                                         na.rm = TRUE)]
  
  if ( eStringLoc < n )
    msg = c(msg, body[ (eStringLoc + 1) : n ])
  
  if ( !is.null(attachLocs) ) {
    attachLens = diff(attachLocs, lag = 1) 
    attachTypes = mapply(function(begL, endL) {
      contentTypeLoc = grep("[Cc]ontent-[Tt]ype", body[ (begL + 1) : (endL - 1)])
      contentType = body[ begL + contentTypeLoc]
      contentType = gsub('"', "", contentType )
      MIMEType = sub(" *Content-Type: *([^;]*);?.*", "\\1", contentType)
      return(MIMEType)
    }, attachLocs[-length(attachLocs)], attachLocs[-1])
  }
  
  if (is.null(attachLocs)) return(list(body = msg, attachInfo = NULL) )
  else return(list(body = msg, 
                   attachDF = data.frame(aLen = attachLens, 
                                         aType = attachTypes,
                                         stringsAsFactors = FALSE)))                                
}
bodyList = lapply(sampleSplit, function(msg) msg$body)
attList = mapply(processAttach, bodyList[hasAttach], 
                 contentTypes[hasAttach], 
                 SIMPLIFY = FALSE)
lens = sapply(attList, function(processedA) 
                           processedA$attachDF$aLen)
head(lens)  # will cause an error with the first message. its ok.
attList[[2]]$attachDF
body = bodyList[hasAttach][[2]]
length(body)
body[35:45]
```

```{r}
processAttach = function(body, contentType){
  n = length(body)
  boundary = getBoundary(contentType)
 
  bString = paste("--", boundary, sep = "")
  bStringLocs = which(bString == body)
  eString = paste("--", boundary, "--", sep = "")
  eStringLoc = which(eString == body)
  
  if (length(eStringLoc) == 0) eStringLoc = n
  if (length(bStringLocs) <= 1) {
    attachLocs = NULL
    msgLastLine = n
    if (length(bStringLocs) == 0) bStringLocs = 0
  } else {
    attachLocs = c(bStringLocs[ -1 ],  eStringLoc)
    msgLastLine = bStringLocs[2] - 1
  }
  
  msg = body[ (bStringLocs[1] + 1) : msgLastLine] 
  if ( eStringLoc < n )
    msg = c(msg, body[ (eStringLoc + 1) : n ])
  
  if ( !is.null(attachLocs) ) {
    attachLens = diff(attachLocs, lag = 1) 
    attachTypes = mapply(function(begL, endL) {
      CTloc = grep("^[Cc]ontent-[Tt]ype", body[ (begL + 1) : (endL - 1)])
      if ( length(CTloc) == 0 ) {
        MIMEType = NA
      } else {
        CTval = body[ begL + CTloc[1] ]
        CTval = gsub('"', "", CTval )
        MIMEType = sub(" *[Cc]ontent-[Tt]ype: *([^;]*);?.*", "\\1", CTval)   
      }
      return(MIMEType)
    }, attachLocs[-length(attachLocs)], attachLocs[-1])
  }
  
  if (is.null(attachLocs)) return(list(body = msg, attachDF = NULL) )
  return(list(body = msg, 
             attachDF = data.frame(aLen = attachLens, 
                                     aType = unlist(attachTypes),
                                     stringsAsFactors = FALSE)))                                
}                       
readEmail = function(dirName) {
       # retrieve the names of files in directory
  fileNames = list.files(dirName, full.names = TRUE)
       # drop files that are not email
  notEmail = grep("cmds$", fileNames)
  if ( length(notEmail) > 0) fileNames = fileNames[ - notEmail ]
       # read all files in the directory
  lapply(fileNames, readLines, encoding = "latin1")
}
```

```{r}
processAllEmail = function(dirName, isSpam = FALSE)
{
       # read all files in the directory
  messages = readEmail(dirName)
  fileNames = names(messages)
  n = length(messages)
  
       # split header from body
  eSplit = lapply(messages, splitMessage)
  rm(messages)
       # process header as named character vector
  headerList = lapply(eSplit, function(msg) 
                                 processHeader(msg$header))
  
       # extract content-type key
  contentTypes = sapply(headerList, function(header) 
                                       header["Content-Type"])
  
       # extract the body
  bodyList = lapply(eSplit, function(msg) msg$body)
  rm(eSplit)
       # which email have attachments
  hasAttach = grep("^ *multi", tolower(contentTypes))
       # get summary stats for attachments and the shorter body
  attList = mapply(processAttach, bodyList[hasAttach], 
                   contentTypes[hasAttach], SIMPLIFY = FALSE)
  
  bodyList[hasAttach] = lapply(attList, function(attEl) 
                                           attEl$body)
 
  attachInfo = vector("list", length = n )
  attachInfo[ hasAttach ] = lapply(attList, 
                                  function(attEl) attEl$attachDF)
 
       # prepare return structure
  emailList = mapply(function(header, body, attach, isSpam) {
                       list(isSpam = isSpam, header = header, 
                            body = body, attach = attach)
                     },
                     headerList, bodyList, attachInfo, 
                     rep(isSpam, n), SIMPLIFY = FALSE )
  names(emailList) = fileNames
  
  invisible(emailList)
}
emailStruct = mapply(processAllEmail, fullDirNames,
                     isSpam = rep( c(FALSE, TRUE), 3:2))      
emailStruct = unlist(emailStruct, recursive = FALSE)
sampleStruct = emailStruct[ indx ]
save(emailStruct, file="emailXX.rda")
header = sampleStruct[[1]]$header
subject = header["Subject"]
els = strsplit(subject, "")
all(els %in% LETTERS)
testSubject = c("DEAR MADAME", "WINNER!", "")
els = strsplit(testSubject, "")
sapply(els, function(subject) all(subject %in% LETTERS))
gsub("[[:punct:] ]", "", testSubject)
gsub("[^[:alpha:]]", "", testSubject)
isYelling = function(msg) {
  if ( "Subject" %in% names(msg$header) ) {
     el = gsub("[^[:alpha:]]", "", msg$header["Subject"])
     if (nchar(el) > 0) 
        nchar(gsub("[A-Z]", "", el)) < 1
     else 
        FALSE
  } else 
    NA
}
perCaps =
function(msg)
{
  body = paste(msg$body, collapse = "")
       # Return NA if the body of the message is "empty"
  if(length(body) == 0 || nchar(body) == 0) return(NA)
       # Eliminate non-alpha characters
  body = gsub("[^[:alpha:]]", "", body)
  capText = gsub("[^A-Z]", "", body)
  100 * nchar(capText)/nchar(body)
}
sapply(sampleStruct, perCaps)
funcList = list( 
 isRe = function(msg) {
       "Subject" %in% names(msg$header) &&
         length(grep("^[ \t]*Re:", msg$header[["Subject"]])) > 0
  },
  numLines = function(msg) 
                length(msg$body),
  isYelling = function(msg) {
    if ( "Subject" %in% names(msg$header) ) {
       el = gsub("[^[:alpha:]]", "", msg$header["Subject"])
       if (nchar(el) > 0) 
         nchar(gsub("[A-Z]", "", el)) < 1
       else 
         FALSE
    }
    else NA
  },
  perCaps = function(msg) {
    body = paste(msg$body, collapse = "")
         # Return NA if the body of the message is "empty"
    if(length(body) == 0 || nchar(body) == 0) return(NA)
         # Eliminate non-alpha characters
    body = gsub("[^[:alpha:]]", "", body)
    capText = gsub("[^A-Z]", "", body)
    100 * nchar(capText)/nchar(body)
  }
)
lapply(funcList, function(func) 
                   sapply(sampleStruct, function(msg) func(msg)))
createDerivedDF =
function(email = emailStruct, operations = funcList, 
         verbose = FALSE)
{
  els = lapply(names(operations),
               function(id) {
                 if(verbose) print(id)
                 e = operations[[id]]
                 v = if(is.function(e)) 
                        sapply(email, e)
                      else 
                        sapply(email, function(msg) eval(e))
                 v
         })
   df = as.data.frame(els)
   names(df) = names(operations)
   invisible(df)
}
sampleDF = createDerivedDF(sampleStruct)
head(sampleDF)
funcList = list(
  isSpam =
    expression(msg$isSpam)
  ,
  isRe =
    function(msg) {
      # Can have a Fwd: Re:  ... but we are not looking for this here.
      # We may want to look at In-Reply-To field.
      "Subject" %in% names(msg$header) && 
        length(grep("^[ \t]*Re:", msg$header[["Subject"]])) > 0
    }
  ,
  numLines =
    function(msg) length(msg$body)
  ,
  bodyCharCt =
    function(msg)
      sum(nchar(msg$body))
  ,
  underscore =
    function(msg) {
      if(!"Reply-To" %in% names(msg$header))
        return(FALSE)
      
      txt <- msg$header[["Reply-To"]]
      length(grep("_", txt)) > 0  && 
        length(grep("[0-9A-Za-z]+", txt)) > 0
    }
  ,
  subExcCt = 
    function(msg) {
      x = msg$header["Subject"]
      if(length(x) == 0 || sum(nchar(x)) == 0 || is.na(x))
        return(NA)
      
      sum(nchar(gsub("[^!]","", x)))
    }
  ,
  subQuesCt =
    function(msg) {
      x = msg$header["Subject"]
      if(length(x) == 0 || sum(nchar(x)) == 0 || is.na(x))
        return(NA)
      
      sum(nchar(gsub("[^?]","", x)))
    }
  ,
  numAtt = 
    function(msg) {
      if (is.null(msg$attach)) return(0)
      else nrow(msg$attach)
    }
   
  ,
  priority =
    function(msg) {
      ans <- FALSE
      # Look for names X-Priority, Priority, X-Msmail-Priority
      # Look for high any where in the value
      ind = grep("priority", tolower(names(msg$header)))
      if (length(ind) > 0)  {
        ans <- length(grep("high", tolower(msg$header[ind]))) >0
      }
      ans
    }
  ,
  numRec =
    function(msg) {
      # unique or not.
      els = getMessageRecipients(msg$header)
      
      if(length(els) == 0)
        return(NA)
      
      # Split each line by ","  and in each of these elements, look for
      # the @ sign. This handles
      tmp = sapply(strsplit(els, ","), function(x) grep("@", x))
      sum(sapply(tmp, length))
    }
  ,
  perCaps =
    function(msg)
    {
      body = paste(msg$body, collapse = "")
      
      # Return NA if the body of the message is "empty"
      if(length(body) == 0 || nchar(body) == 0) return(NA)
      
      # Eliminate non-alpha characters and empty lines 
      body = gsub("[^[:alpha:]]", "", body)
      els = unlist(strsplit(body, ""))
      ctCap = sum(els %in% LETTERS)
      100 * ctCap / length(els)
    }
  ,
  isInReplyTo =
    function(msg)
    {
      "In-Reply-To" %in% names(msg$header)
    }
  ,
  sortedRec =
    function(msg)
    {
      ids = getMessageRecipients(msg$header)
      all(sort(ids) == ids)
    }
  ,
  subPunc =
    function(msg)
    {
      if("Subject" %in% names(msg$header)) {
        el = gsub("['/.:@-]", "", msg$header["Subject"])
        length(grep("[A-Za-z][[:punct:]]+[A-Za-z]", el)) > 0
      }
      else
        FALSE
    },
  hour =
    function(msg)
    {
      date = msg$header["Date"]
      if ( is.null(date) ) return(NA)
      # Need to handle that there may be only one digit in the hour
      locate = regexpr("[0-2]?[0-9]:[0-5][0-9]:[0-5][0-9]", date)
      
      if (locate < 0)
        locate = regexpr("[0-2]?[0-9]:[0-5][0-9]", date)
      if (locate < 0) return(NA)
      
      hour = substring(date, locate, locate+1)
      hour = as.numeric(gsub(":", "", hour))
      
      locate = regexpr("PM", date)
      if (locate > 0) hour = hour + 12
      
      locate = regexpr("[+-][0-2][0-9]00", date)
      if (locate < 0) offset = 0
      else offset = as.numeric(substring(date, locate, locate + 2))
      (hour - offset) %% 24
    }
  ,
  multipartText =
    function(msg)
    {
      if (is.null(msg$attach)) return(FALSE)
      numAtt = nrow(msg$attach)
      
      types = 
        length(grep("(html|plain|text)", msg$attach$aType)) > (numAtt/2)
    }
  ,
  hasImages =
    function(msg)
    {
      if (is.null(msg$attach)) return(FALSE)
      
      length(grep("^ *image", tolower(msg$attach$aType))) > 0
    }
  ,
  isPGPsigned =
    function(msg)
    {
      if (is.null(msg$attach)) return(FALSE)
      
      length(grep("pgp", tolower(msg$attach$aType))) > 0
    },
  perHTML =
    function(msg)
    {
      if(! ("Content-Type" %in% names(msg$header))) return(0)
      
      el = tolower(msg$header["Content-Type"]) 
      if (length(grep("html", el)) == 0) return(0)
      
      els = gsub("[[:space:]]", "", msg$body)
      totchar = sum(nchar(els))
      totplain = sum(nchar(gsub("<[^<]+>", "", els )))
      100 * (totchar - totplain)/totchar
    },
  subSpamWords =
    function(msg)
    {
      if("Subject" %in% names(msg$header))
        length(grep(paste(SpamCheckWords, collapse = "|"), 
                    tolower(msg$header["Subject"]))) > 0
      else
        NA
    }
  ,
  subBlanks =
    function(msg)
    {
      if("Subject" %in% names(msg$header)) {
        x = msg$header["Subject"]
        # should we count blank subject line as 0 or 1 or NA?
        if (nchar(x) == 1) return(0)
        else 100 *(1 - (nchar(gsub("[[:blank:]]", "", x))/nchar(x)))
      } else NA
    }
  ,
  noHost =
    function(msg)
    {
      # Or use partial matching.
      idx = pmatch("Message-", names(msg$header))
      
      if(is.na(idx)) return(NA)
      
      tmp = msg$header[idx]
      return(length(grep(".*@[^[:space:]]+", tmp)) ==  0)
    }
  ,
  numEnd =
    function(msg)
    {
      # If we just do a grep("[0-9]@",  )
      # we get matches on messages that have a From something like
      # " \"marty66@aol.com\" <synjan@ecis.com>"
      # and the marty66 is the "user's name" not the login
      # So we can be more precise if we want.
      x = names(msg$header)
      if ( !( "From" %in% x) ) return(NA)
      login = gsub("^.*<", "", msg$header["From"])
      if ( is.null(login) ) 
        login = gsub("^.*<", "", msg$header["X-From"])
      if ( is.null(login) ) return(NA)
      login = strsplit(login, "@")[[1]][1]
      length(grep("[0-9]+$", login)) > 0
    },
  isYelling =
    function(msg)
    {
      if ( "Subject" %in% names(msg$header) ) {
        el = gsub("[^[:alpha:]]", "", msg$header["Subject"])
        if (nchar(el) > 0) nchar(gsub("[A-Z]", "", el)) < 1
        else FALSE
      }
      else
        NA
    },
  forwards =
    function(msg)
    {
      x = msg$body
      if(length(x) == 0 || sum(nchar(x)) == 0)
        return(NA)
      
      ans = length(grep("^[[:space:]]*>", x))
      100 * ans / length(x)
    },
  isOrigMsg =
    function(msg)
    {
      x = msg$body
      if(length(x) == 0) return(NA)
      
      length(grep("^[^[:alpha:]]*original[^[:alpha:]]+message[^[:alpha:]]*$", 
                  tolower(x) ) ) > 0
    },
  isDear =
    function(msg)
    {
      x = msg$body
      if(length(x) == 0) return(NA)
      
      length(grep("^[[:blank:]]*dear +(sir|madam)\\>", 
                  tolower(x))) > 0
    },
  isWrote =
    function(msg)
    {
      x = msg$body
      if(length(x) == 0) return(NA)
      
      length(grep("(wrote|schrieb|ecrit|escribe):", tolower(x) )) > 0
    },
  avgWordLen =
    function(msg)
    {
      txt = paste(msg$body, collapse = " ")
      if(length(txt) == 0 || sum(nchar(txt)) == 0) return(0)
      
      txt = gsub("[^[:alpha:]]", " ", txt)
      words = unlist(strsplit(txt, "[[:blank:]]+"))
      wordLens = nchar(words)
      mean(wordLens[ wordLens > 0 ])
    }
  ,
  numDlr =
    function(msg)
    {
      x = paste(msg$body, collapse = "")
      if(length(x) == 0 || sum(nchar(x)) == 0)
        return(NA)
      
      nchar(gsub("[^$]","", x))
    }
)
SpamCheckWords =
  c("viagra", "pounds", "free", "weight", "guarantee", "million", 
    "dollars", "credit", "risk", "prescription", "generic", "drug",
    "financial", "save", "dollar", "erotic", "million", "barrister",
    "beneficiary", "easy", 
    "money back", "money", "credit card")
getMessageRecipients =
  function(header)
  {
    c(if("To" %in% names(header))  header[["To"]] else character(0),
      if("Cc" %in% names(header))  header[["Cc"]] else character(0),
      if("Bcc" %in% names(header)) header[["Bcc"]] else character(0)
    )
  }
emailDF = createDerivedDF(emailStruct)
dim(emailDF)
#save(emailDF, file = "spamAssassinDerivedDF.rda")
#load("Data/spamAssassinDerivedDF.rda")
dim(emailDF)
perCaps2 =
function(msg)
{
  body = paste(msg$body, collapse = "")
       # Return NA if the body of the message is "empty"
  if(length(body) == 0 || nchar(body) == 0) return(NA)
       # Eliminate non-alpha characters and empty lines 
  body = gsub("[^[:alpha:]]", "", body)
  els = unlist(strsplit(body, ""))
  ctCap = sum(els %in% LETTERS)
  100 * ctCap / length(els)
}
pC = sapply(emailStruct, perCaps)
pC2 = sapply(emailStruct, perCaps2)
identical(pC, pC2)
indNA = which(is.na(emailDF$subExcCt))
indNoSubject = which(sapply(emailStruct, 
                            function(msg) 
                              !("Subject" %in% names(msg$header))))
all(indNA == indNoSubject)
all(emailDF$bodyCharCt > emailDF$numLines)
```

```{r}
x.at = c(1,10,100,1000,10000,100000)
y.at = c(1, 5, 10, 50, 100, 500, 5000)
nL = 1 + emailDF$numLines
nC = 1 + emailDF$bodyCharCt
#pdf("ScatterPlotNumLinesNumChars.pdf", width = 6, height = 4.5)
plot(nL ~ nC, log = "xy", pch=".", xlim=c(1,100000), axes = FALSE,
     xlab = "Number of Characters", ylab = "Number of Lines")
box() 
axis(1, at = x.at, labels = formatC(x.at, digits = 0, format="d"))
axis(2, at = y.at, labels = formatC(y.at, digits = 0, format="d")) 
abline(a=0, b=1, col="red", lwd = 2)
dev.off()
#pdf("SPAM_boxplotsPercentCaps.pdf", width = 5, height = 5)
percent = emailDF$perCaps
isSpamLabs = factor(emailDF$isSpam, labels = c("ham", "spam"))
boxplot(log(1 + percent) ~ isSpamLabs,
        ylab = "Percent Capitals (log)")
#dev.off()
```

```{r}
logPerCapsSpam = log(1 + emailDF$perCaps[ emailDF$isSpam ])
logPerCapsHam = log(1 + emailDF$perCaps[ !emailDF$isSpam ])
qqplot(logPerCapsSpam, logPerCapsHam, 
       xlab = "Regular Email", ylab = "Spam Email", 
       main = "Percentage of Capital Letters (log scale)",
       pch = 19, cex = 0.3)
```

```{r}
#pdf("SPAM_scatterplotPercentCapsTotChars.pdf", width = 8, height = 6)
 
colI = c("#4DAF4A80", "#984EA380")
logBodyCharCt = log(1 + emailDF$bodyCharCt)
logPerCaps = log(1 + emailDF$perCaps)
plot(logPerCaps ~ logBodyCharCt, xlab = "Total Characters (log)",
     ylab = "Percent Capitals (log)",
     col = colI[1 + emailDF$isSpam],
     xlim = c(2,12), pch = 19, cex = 0.5)
#dev.off()
```

```{r}
table(emailDF$numAtt, isSpamLabs)
#pdf("SPAM_mosaicPlots.pdf", width = 8, height = 4)
```

```{r}
oldPar = par(mfrow = c(1, 2), mar = c(1,1,1,1))
colM = c("#E41A1C80", "#377EB880")
isRe = factor(emailDF$isRe, labels = c("no Re:", "Re:"))
mosaicplot(table(isSpamLabs, isRe), main = "",
           xlab = "", ylab = "", color = colM)
fromNE = factor(emailDF$numEnd, labels = c("No #", "#"))
mosaicplot(table(isSpamLabs, fromNE), color = colM,
           main = "", xlab="", ylab = "")
par(oldPar)
#dev.off()
```

```{r}
library(rpart)
setupRpart = function(data) {
  logicalVars = which(sapply(data, is.logical))
  facVars = lapply(data[ , logicalVars], 
                   function(x) {
                      x = as.factor(x)
                      levels(x) = c("F", "T")
                      x
                   })
  cbind(facVars, data[ , - logicalVars])
}
emailDFrp = setupRpart(emailDF)
set.seed(418910)
testSpamIdx = sample(numSpam, size = floor(numSpam/3))
testHamIdx = sample(numHam, size = floor(numHam/3))
testDF = 
  rbind( emailDFrp[ emailDFrp$isSpam == "T", ][testSpamIdx, ],
         emailDFrp[emailDFrp$isSpam == "F", ][testHamIdx, ] )
trainDF =
  rbind( emailDFrp[emailDFrp$isSpam == "T", ][-testSpamIdx, ], 
         emailDFrp[emailDFrp$isSpam == "F", ][-testHamIdx, ])
rpartFit = rpart(isSpam ~ ., data = trainDF, method = "class")
```

```{r}
#install.packages("rpart.plot")
library(rpart.plot)
prp(rpartFit, extra = 1)
library(rpart.plot)
#pdf("SPAM_rpartTree.pdf", width = 7, height = 7)
prp(rpartFit, extra = 1)
#dev.off()
predictions = predict(rpartFit, 
       newdata = testDF[, names(testDF) != "isSpam"],
       type = "class")
 
predsForHam = predictions[ testDF$isSpam == "F" ]
summary(predsForHam)
sum(predsForHam == "T") / length(predsForHam)
predsForSpam = predictions[ testDF$isSpam == "T" ]
sum(predsForSpam == "F") / length(predsForSpam)
complexityVals = c(seq(0.00001, 0.0001, length=19),
                   seq(0.0001, 0.001, length=19), 
                   seq(0.001, 0.005, length=9),
                   seq(0.005, 0.01, length=9))
fits = lapply(complexityVals, function(x) {
         rpartObj = rpart(isSpam ~ ., data = trainDF,
                          method="class", 
                          control = rpart.control(cp=x) )
           
         predict(rpartObj, 
                 newdata = testDF[ , names(testDF) != "isSpam"],
                 type = "class")
        })
spam = testDF$isSpam == "T"
numSpam = sum(spam)
numHam = sum(!spam)
errs = sapply(fits, function(preds) {
                      typeI = sum(preds[ !spam ] == "T") / numHam
                      typeII = sum(preds[ spam ] == "F") / numSpam
                      c(typeI = typeI, typeII = typeII)
                     })
```

```{r}
#pdf("SPAM_rpartTypeIandII.pdf", width = 8, height = 7)
library(RColorBrewer)
cols = brewer.pal(9, "Set1")[c(3, 4, 5)]
plot(errs[1,] ~ complexityVals, type="l", col=cols[2], 
     lwd = 2, ylim = c(0,0.2), xlim = c(0,0.01), 
     ylab="Error", xlab="complexity parameter values")
points(errs[2,] ~ complexityVals, type="l", col=cols[1], lwd = 2)
text(x =c(0.003, 0.0035), y = c(0.12, 0.05), 
     labels=c("Type II Error", "Type I Error"))
minI = which(errs[1,] == min(errs[1,]))[1]
abline(v = complexityVals[minI], col ="grey", lty =3, lwd=2)
text(0.0007, errs[1, minI]+0.01, 
     formatC(errs[1, minI], digits = 2))
text(0.0007, errs[2, minI]+0.01, 
     formatC(errs[2, minI], digits = 3))
#dev.off()
save(emailDFrp,file="data.Rda")
```

```{r}
library(caret)
setupRnum = function(data) {
  logicalVars = which(sapply(data, is.logical))
  facVars = lapply(data[ , logicalVars], 
                   function(x) {
                      x = as.numeric(x)
                   })
  cbind(facVars, data[ , - logicalVars])
}
emailDFnum = setupRnum(emailDF)
emailDFnum[is.na(emailDFnum)]<-0
cv_folds <- createFolds(emailDFnum$isSpam, k=5, list=TRUE, returnTrain = TRUE)
lengths(cv_folds)
```

```{r}
library(MLmetrics)
f1 <- function(data, lev = NULL, model = NULL) {
  f1_val <- F1_Score(y_pred = data$pred, y_true = data$obs, positive = lev[1])
  p <- Precision(y_pred = data$pred, y_true = data$obs, positive = lev[1])
  r <- Recall(y_pred = data$pred, y_true = data$obs, positive = lev[1])
  fp <-sum(data$pred==0 & data$obs==1)/length(data$pred)  
 
  fn <-sum(data$pred==1 & data$obs==0)/length(data$pred)
    c(F1 = f1_val,
    prec = p,
    rec = r,
    Type_I_err=fp,
    Type_II_err=fn
   )
}
```

```{r}
library(naivebayes)
library(e1071)
nb_grid<-expand.grid(laplace=c(0,0.1,0.3,0.5,1), usekernel=c(T,F), adjust=c(T,F))
train_control<-trainControl(method="cv", number=3, savePredictions = 'final',summaryFunction = f1)
model_nb<-caret::train(as.factor(isSpam) ~ .,data=emailDFnum, trControl = train_control, method='naive_bayes',tuneGrid = nb_grid)
model_nb
#Did the boss fool us with the folds?  Nope.
table(model_nb$pred['Resample'])
val<-seq(from = 0, to=0.01, by=0.0005)
```

```{r}
library(rpart)
cart_grid<-expand.grid(cp=val)
train_control<-trainControl(method="cv", number =5, savePredictions = 'final',summaryFunction = f1)
model_rpart<-caret::train(as.factor(isSpam) ~ .,data=emailDFnum, trControl = train_control, method='rpart',tuneGrid = cart_grid)
model_rpart
```

```{r}
library(randomForest)
rf_grid<-expand.grid(mtry=seq(from =1, to = 25, by = 2))
train_control<-trainControl(method="cv", number=3, savePredictions = 'final',summaryFunction = f1)
model_rf<-caret::train(as.factor(isSpam) ~ .,data=emailDFnum, trControl = train_control, ntree=200,method='rf',tuneGrid = rf_grid)
model_rf
```

```{r}
library(xgboost)
xgb_grid<-expand.grid(nrounds = 100, max_depth = c(3,5,7,9,11), eta = c(0.01,0.03,0.1), gamma=c(1,3,5,10), colsample_bytree=1, min_child_weight=1, subsample=1)
train_control<-trainControl(method="cv", number=3, savePredictions = 'final',summaryFunction = f1)
model_xgb<-caret::train(as.factor(isSpam) ~ .,data=emailDFnum, trControl = train_control,method='xgbTree',tuneGrid = xgb_grid)